---
title: "Getting started with dqcheckr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dqcheckr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

`dqcheckr` automates the verification of recurring external dataset deliveries.
For each new file arrival it runs a battery of quality checks, compares the
file to the previous delivery, writes a self-contained HTML report, and records
summary statistics in a local SQLite database so that quality trends can be
tracked over time.

## How it works

```
Your data file                  Your YAML config
      │                                │
      ▼                                ▼
  read_dataset()            load_config()
  (all columns as           (global defaults +
   character, trimmed)       per-dataset rules)
      │                                │
      └──────────────┬─────────────────┘
                     ▼
          run_qc_checks()        ← single-snapshot checks (QC / SC series)
          run_comparison_checks()← vs previous delivery   (CP series)
          run_custom_checks()    ← your own rules
                     │
                     ▼
          write_snapshot()       ← SQLite: run summary + per-column stats
          render_report()        ← self-contained HTML
                     │
                     ▼
          run_dq_check()         ← the one function you call
```

A data officer runs a single command for each arriving dataset:

```{r one-call}
run_dq_check("customer_accounts", config_dir = "path/to/configs")
```

This prints a one-line console summary, writes an HTML report, and returns
`list(status, report_path, snapshot_id)` invisibly.

---

## Installation

`dqcheckr` is a first-party package — install it directly from source:

```{r install}
devtools::install("path/to/dqr/dqcheckr")
library(dqcheckr)
```

All dependencies are on CRAN: `readr`, `DBI`, `RSQLite`, `rmarkdown`,
`knitr`, `kableExtra`, `ggplot2`, `gridExtra`, `yaml`, `dplyr`, `tidyr`,
`rlang`.

---

## Configuration

Two YAML files control every run.

### Global config — `dqcheckr.yml`

Placed in your config directory. Sets default thresholds used by all datasets
unless a dataset overrides them.

```yaml
snapshot_db:       "data/snapshots.sqlite"
report_output_dir: "reports/"

default_rules:
  # Proportion of non-empty values that must parse as numeric for a column to
  # be classified as numeric (affects type inference across many checks)
  type_inference_threshold: 0.90

  # FAIL if any column's missing-value rate exceeds this (0-1)
  max_missing_rate: 0.05

  # FAIL if non-numeric values in a numeric column exceed this rate
  max_non_numeric_rate: 0.01

  # FAIL if row count drops below this (0 = disabled)
  min_row_count: 0

  # Version-comparison thresholds (used when a previous file is available)
  max_row_count_change_pct:       0.10
  max_numeric_mean_shift_pct:     0.20
  max_missing_rate_change_pp:     2.0    # percentage points
  max_non_numeric_rate_change_pp: 1.0

  flag_new_columns:         true   # set false to suppress new-column warnings in CP-02
  flag_dropped_columns:     true   # set false to suppress dropped-column warnings in CP-02
  flag_type_changes:        true   # set false to suppress type-change warnings in CP-02
  flag_column_order_change: true   # set false to skip CP-08 entirely (WARN CSV / FAIL FWF)
```

### Per-dataset config — `<dataset_name>.yml`

One file per dataset. Only the keys you specify are needed — everything else
inherits from the global defaults.

```yaml
dataset_name: "customer_accounts"   # for human reference only — the functional
                                     # identifier is the YAML filename, which must
                                     # match the dataset_name argument to run_dq_check()

# --- File location ---
# Option A: folder scan (two most recently modified files are used)
folder: "data/incoming/customer_accounts/"

# Option B: explicit paths (set current_file only → single-file mode,
#           no version comparison)
# current_file:  "data/incoming/customer_accounts/20260410.csv"
# previous_file: "data/incoming/customer_accounts/20260401.csv"

format:    csv
encoding:  "UTF-8"
delimiter: ","

# --- Schema contract (optional) ---
# SC-01 flags extra columns; SC-02 flags missing columns.
expected_columns:
  - id
  - name
  - country_code
  - account_status
  - account_balance
  - created_date

# --- Key columns (optional) ---
# QC-12 checks each for duplicate values.
key_columns:
  - id

# --- Rule overrides (optional) ---
# These replace the matching keys in default_rules for this dataset only.
rule_overrides:
  max_missing_rate:        0.02
  max_row_count_change_pct: 0.05
  min_row_count:           1000

# --- Per-column type overrides (optional) ---
# Force specific columns to a type regardless of what the data looks like.
# Valid values: character, numeric, date.
# Eliminates false positives on phone numbers, postcodes, unit numbers, etc.
column_types:
  phone:    character   # prevent QC-11 / CP-04 / CP-07 false positives
  postcode: character   # 4-digit codes are not meaningful as numeric
  bsb_code: character   # 6-digit bank codes are not meaningful as numeric

# --- Per-column rules (optional) ---
column_rules:
  country_code:
    allowed_values: ["GB", "US", "DE", "FR"]   # QC-09
    pattern: "^[A-Z]{2}$"                       # QC-13
  account_status:
    allowed_values: ["ACTIVE", "CLOSED", "SUSPENDED"]
  account_balance:
    min_value: 0                                 # QC-10
    max_value: 1000000
  # Per-column threshold overrides (optional, within column_rules):
  email:
    max_missing_rate: 1.00          # email is 99% absent by design — suppress QC-01
    max_missing_rate_change_pp: 5.0 # looser than global 2.0 — suppress CP-03 noise
  account_balance:
    max_missing_rate: 0.00          # balance must always be present
    max_numeric_mean_shift_pct: 0.05  # stricter than global 0.20

# --- Custom checks (optional) ---
# custom_checks_file: "custom/customer_accounts_checks.R"

# --- Output paths (optional, override global) ---
# snapshot_db:       "data/snapshots.sqlite"
# report_output_dir: "reports/"
```

### What is required vs optional

The only fields the package requires to run are the file location (`folder`
or `current_file`) and the `format`. Every other config key is optional —
the corresponding checks are simply skipped when the key is absent.

| Config key | Skipped / default when absent |
|---|---|
| `key_columns` | QC-12 (key uniqueness) skipped |
| `expected_columns` | SC-01 and SC-02 (schema contract) skipped |
| `column_types` | All columns use automatic type inference |
| `column_rules.allowed_values` | QC-09 (allowed values) skipped |
| `column_rules.min_value` / `max_value` | QC-10 (numeric bounds) skipped |
| `column_rules.pattern` | QC-13 (pattern / regex) skipped |
| `column_rules.<col>.max_missing_rate` | Falls back to dataset / global `max_missing_rate` |
| `column_rules.<col>.max_non_numeric_rate` | Falls back to dataset / global `max_non_numeric_rate` |
| `column_rules.<col>.max_missing_rate_change_pp` | Falls back to dataset / global `max_missing_rate_change_pp` |
| `column_rules.<col>.max_numeric_mean_shift_pct` | Falls back to dataset / global `max_numeric_mean_shift_pct` |
| `column_rules.<col>.max_non_numeric_rate_change_pp` | Falls back to dataset / global `max_non_numeric_rate_change_pp` |
| `custom_checks_file` | All custom checks skipped |
| `previous_file` / second file in folder | All CP-01–CP-08 (version comparison) skipped |
| `min_row_count` set to 0 | QC-14 (minimum row count) skipped |
| `type_inference_threshold` | Defaults to 0.90 (affects QC-06, QC-07, QC-08, QC-11, CP-02, CP-04, CP-05, CP-06, CP-07) |
| `flag_new_columns` | Defaults to `true`; set to `false` to suppress new-column warnings in CP-02 (still tracked in SQLite) |
| `flag_dropped_columns` | Defaults to `true`; set to `false` to suppress dropped-column warnings in CP-02 (still tracked in SQLite) |
| `flag_type_changes` | Defaults to `true`; set to `false` to suppress type-change warnings in CP-02 |
| `flag_column_order_change` | Defaults to `true`; set to `false` to skip CP-08 entirely |

A minimal dataset config that runs basic checks with no column-level rules:

```yaml
dataset_name: "my_dataset"           # for reference; must match the YAML filename
current_file: "data/incoming/my_dataset.csv"
format: csv
```

### Fixed-width files

For fixed-width files, set `format: fwf` and supply column widths and names.
If the file has a header row, set `fwf_skip: 1` to skip it.

```yaml
format: fwf
fwf_skip: 1
fwf_widths:    [10, 20, 3, 10, 12, 11]
fwf_col_names: [id, name, country_code, account_status, account_balance, created_date]
```

---

## The quality checks

Each check returns one or more `dq_result` objects with four key fields:
`check_id`, `status` (`PASS` / `WARN` / `FAIL` / `INFO`), `observed`, and
`message`. The overall run status is the worst single status across all checks.

### Single-snapshot checks (QC series)

These run on the current file alone.

| ID | Name | Triggers on | Severity |
|----|------|-------------|----------|
| QC-01 | Missing rate | Column missing rate > `max_missing_rate` | FAIL |
| QC-02 | Empty column | Column is 100% empty | FAIL |
| QC-03 | Duplicate rows | Any fully-identical rows | WARN |
| QC-04 | Row count | Always | INFO |
| QC-05 | Column count | Always | INFO |
| QC-06 | Inferred type | Always (date / numeric / character / unknown) | INFO |
| QC-07 | Numeric stats | Numeric columns (min, max, mean, SD) | INFO |
| QC-08 | Distinct count | Character columns | INFO |
| QC-09 | Allowed values | Values outside `allowed_values` list | FAIL |
| QC-10 | Numeric bounds | Values outside `min_value` / `max_value` | FAIL |
| QC-11 | Non-numeric values | Non-numeric values in a numeric column | WARN / FAIL |
| QC-12 | Key uniqueness | Duplicate values in a `key_columns` column | FAIL |
| QC-13 | Pattern / regex | Values not matching `pattern` | FAIL |
| QC-14 | Minimum row count | Row count below `min_row_count` | FAIL |

### Schema contract checks (SC series)

Only run when `expected_columns` is set.

| ID | Name | Triggers on | Severity |
|----|------|-------------|----------|
| SC-01 | Unexpected column | Column in file not in `expected_columns` | FAIL |
| SC-02 | Missing expected column | Column in `expected_columns` absent from file | FAIL |

### Version comparison checks (CP series)

Only run when a previous file is available. Skipped automatically in
single-file mode.

| ID | Name | Triggers on | Severity |
|----|------|-------------|----------|
| CP-01 | Row count change | Change > `max_row_count_change_pct` | WARN |
| CP-02 | Schema diff | Columns added, dropped, or type-changed | WARN |
| CP-03 | Missing rate change | Increase > `max_missing_rate_change_pp` pp | WARN |
| CP-04 | Numeric mean shift | Shift > `max_numeric_mean_shift_pct` | WARN |
| CP-05 | New distinct values | Values in current not seen in previous | INFO |
| CP-06 | Dropped distinct values | Values in previous absent from current | INFO |
| CP-07 | Non-numeric rate change | Increase > `max_non_numeric_rate_change_pp` pp | WARN |
| CP-08 | Column order change | Column order differs from previous | WARN (CSV) / FAIL (FWF) |

---

## Type inference

`infer_col_type()` classifies each column after whitespace trimming. Rules
are applied in priority order:

1. **date** — all non-empty values parse as a known date format
   (`%Y-%m-%d`, `%d/%m/%Y`, `%m/%d/%Y`, `%Y%m%d`, `%d-%m-%Y`)
2. **numeric** — ≥ `type_inference_threshold` of non-empty values coerce to numeric
3. **character** — everything else
4. **unknown** — all values are empty or `NA`

The threshold defaults to 90%, meaning a column with up to 10% bad values is
still classified as numeric, making those bad values visible to QC-11
(non-numeric contamination). It is configurable per dataset via
`type_inference_threshold` in `default_rules` or `rule_overrides`:

```yaml
# raise threshold — stricter, fewer columns classified as numeric
rule_overrides:
  type_inference_threshold: 0.95

# lower threshold — more lenient, columns with more noise still classified numeric
rule_overrides:
  type_inference_threshold: 0.80
```

Changing the threshold affects QC-06, QC-07, QC-08, QC-11, CP-02, CP-04,
CP-05, CP-06, and CP-07. Note that changing it between runs may cause CP-02
to report a spurious type change on columns that sit close to the old and new
thresholds, even if the underlying data did not change.

```{r type-inference}
infer_col_type(c("2024-01-15", "2024-06-30"))  # "date"
infer_col_type(c("100", "200", "N/A"))          # "character" (only 67% numeric)
infer_col_type(c("100", "200", "N/A", rep("50", 17)))  # "numeric" (95% numeric)
infer_col_type(c(NA, "", NA))                   # "unknown"
# with a custom threshold:
infer_col_type(c(rep("1", 17), "a", "b", "c"), threshold = 0.80)  # "numeric" (85%)
```

### Overriding inferred types

Some columns are formatted as numbers but are semantically character — phone
numbers, postcodes, unit numbers, BSB codes. Automatic inference classifies
them as `numeric`, which triggers irrelevant QC-11, CP-04, and CP-07 checks
and can cause spurious CP-02 type-change warnings when the proportion of
non-numeric values drifts across the inference threshold between deliveries.

Use `column_types` in the dataset YAML to force the type for any column:

```yaml
column_types:
  TenantPhone:    character   # 10-digit numbers — not meaningful as numeric
  PremisesUnit:   character   # values like "5", "3A", "Ground Floor"
  PremisesPostCode: character # 4-digit codes validated by pattern, not numeric
  BSBCode:        character   # 6-digit bank routing codes
```

Valid values are `character`, `numeric`, and `date`. The override affects all
downstream logic: the type is stored in SQLite as the column's `inferred_type`,
drift comparisons see the stable override, and QC/CP checks that would be
meaningless for the forced type are skipped automatically.

Within custom check scripts, declare `config` as a second argument and use
`resolve_col_type()` instead of `infer_col_type()` so that overrides are
respected. If the function signature has only one argument (`df`), the config
is not passed — both signatures work:

```{r resolve-col-type}
# Two-argument form — config is passed automatically, overrides respected
custom_checks <- function(df, config) {
  results <- list()
  for (col in names(df)) {
    if (resolve_col_type(col, df[[col]], config) == "numeric") {
      # numeric-specific check logic here
    }
  }
  results
}

# One-argument form — still works, type inference only (no override awareness)
custom_checks <- function(df) {
  # ...
}
```

---

## Per-column threshold overrides

Global and dataset-level thresholds apply uniformly to all columns. Some
columns naturally have higher missing rates or more variability, making a
single threshold produce constant noise. Per-column threshold keys in
`column_rules` let you tune specific columns without affecting others.

**Resolution order** for every threshold-based check:

```
column_rules.<col>.<threshold>  →  rule_overrides.<threshold>  →  default_rules.<threshold>
```

### Supported per-column threshold keys

| Key | Governs | Default |
|---|---|---|
| `max_missing_rate` | QC-01 (missing rate) | 0.05 |
| `max_non_numeric_rate` | QC-11 (non-numeric contamination) | 0.01 |
| `max_missing_rate_change_pp` | CP-03 (missing rate change) | 2.0 pp |
| `max_numeric_mean_shift_pct` | CP-04 (numeric mean shift) | 0.20 |
| `max_non_numeric_rate_change_pp` | CP-07 (non-numeric rate change) | 1.0 pp |

### Example

```yaml
column_rules:
  # Email is almost always absent by design — don't alert on it
  LandlordEmail:
    max_missing_rate:          1.00   # effectively disables QC-01 for this column
    max_missing_rate_change_pp: 5.0   # allow larger swings without CP-03 noise

  # Bond amount must always be present and should not shift much
  BondAmount:
    max_missing_rate:          0.00   # zero tolerance
    max_numeric_mean_shift_pct: 0.05  # tighter than global 0.20

  # Phone is forced to character via column_types, so numeric thresholds don't apply,
  # but you can still tighten missing-rate expectations:
  TenantPhone:
    max_missing_rate: 0.10
```

The `threshold` field in the HTML report and `dq_result` objects reflects
the effective per-column threshold, not the global default, so the report
always shows the value that was actually used.

---

## Running a check

```{r run-check}
result <- run_dq_check(
  dataset_name = "customer_accounts",
  config_dir   = "path/to/configs",   # contains dqcheckr.yml + customer_accounts.yml
  open_report  = TRUE                 # open HTML in browser when interactive
)

# Console output (always printed):
# [dqcheckr] customer_accounts: FAIL - 1 warning(s), 2 failure(s). Report: reports/...html

result$status       # "PASS", "WARN", or "FAIL"
result$report_path  # full path to the HTML file
result$snapshot_id  # integer row ID in snapshots table
```

The HTML report contains:

1. **Header** — dataset name, file name, run time, overall status badge
2. **File summary** — row count, column count, file size, format, encoding
3. **Quality checks table** — every check result, failures first
4. **Custom checks table** — if configured
5. **Version comparison** — what changed since the previous delivery
6. **Historical trend** — row count and check outcomes across the last 10 runs
   (appears once at least two runs have been recorded)
7. **Appendix** — full column-level statistics table

---

## Calling individual checks

You can call any check function directly, without going through the full
pipeline. Every check takes `(df, config)` and returns a list of `dq_result`
objects.

```{r individual-checks}
library(dqcheckr)

# Build a minimal config programmatically
cfg <- list(
  rules = list(
    max_missing_rate     = 0.05,
    max_non_numeric_rate = 0.01,
    min_row_count        = 0
  ),
  column_rules     = list(
    status = list(allowed_values = c("ACTIVE", "CLOSED"))
  ),
  key_columns      = "id",
  expected_columns = NULL
)

df <- read.csv("data/customer_accounts.csv", colClasses = "character")
df[] <- lapply(df, trimws)

# Run a single check
results <- check_missing_rate(df, cfg)

# Inspect
results[[1]]$status    # "PASS" or "FAIL"
results[[1]]$observed  # e.g. "3.2% missing (4 of 125)"

# Run all single-snapshot checks at once
all_results <- run_qc_checks(df, cfg)

# Overall status
overall_status(all_results)  # "FAIL" > "WARN" > "PASS" > "INFO"
```

---

## Custom checks

Create a plain `.R` file that defines `custom_checks(df)`. It receives the
trimmed data frame and must return a list of `dq_result` objects.

```{r custom-file}
# File: custom/customer_accounts_checks.R

custom_checks <- function(df) {
  results <- list()

  # Rule: ACTIVE accounts must not have a zero balance
  active_zero <- df[df$account_status == "ACTIVE" &
                    !is.na(df$account_balance) &
                    df$account_balance == "0", ]
  n <- nrow(active_zero)
  results <- c(results, list(dq_result(
    check_id   = "CUST-01",
    check_name = "No zero-balance active accounts",
    column     = "account_balance",   # enables per-column storage in SQLite
    status     = if (n > 0) "FAIL" else "PASS",
    observed   = sprintf("%d ACTIVE account(s) with balance 0", n),
    message    = if (n > 0)
      sprintf("%d ACTIVE account(s) have a zero balance.", n)
    else
      "No ACTIVE accounts have a zero balance."
  )))

  results
}
```

Point to it in the dataset config:

```yaml
custom_checks_file: "custom/customer_accounts_checks.R"
```

The file is sourced in an isolated environment (not the global session).
`dq_result()` is available inside it. Any error — missing file, undefined
function, runtime failure — stops the run with a clear message.

**Tip**: always set `column =` when your check is scoped to a specific column.
Results with `column = NA` (the default) appear in the HTML report and
contribute to the overall status, but are not written to the `column_snapshots`
SQLite table and therefore do not appear in per-column trend queries.

---

## The snapshot database

Every run writes two rows to a shared SQLite database: one row in `snapshots`
(run-level summary) and one row per column-stat in `column_snapshots`.

```{r snapshot-query}
library(DBI)
library(RSQLite)

con <- dbConnect(SQLite(), "data/snapshots.sqlite")

# Recent runs for one dataset
dbGetQuery(con,
  "SELECT id, file_name, overall_status, check_fail_count, run_timestamp
   FROM snapshots
   WHERE dataset_name = 'customer_accounts'
   ORDER BY id DESC
   LIMIT 10")

# Column-level stats for the most recent run
dbGetQuery(con,
  "SELECT column_name, dq_check, value, threshold
   FROM column_snapshots
   WHERE snapshot_id = (
     SELECT MAX(id) FROM snapshots WHERE dataset_name = 'customer_accounts'
   )
   ORDER BY column_name, dq_check")

dbDisconnect(con)
```

Because all datasets share one database, you can query across them:

```{r cross-dataset}
dbGetQuery(con,
  "SELECT dataset_name, COUNT(*) AS runs,
          SUM(check_fail_count) AS total_failures
   FROM snapshots
   GROUP BY dataset_name")
```

The `output/` directory (and database file) are created automatically on the
first run if they do not exist.

---

## Historical drift comparison

`run_dq_check()` compares the current file to the previous file in a single
run. For longer-range comparisons — "how has this dataset's quality changed
over the past two years?" — use `compare_snapshots()`, which reads two
historical snapshots from SQLite without needing the original files.

### List available snapshots

```{r list-snapshots}
# All datasets
list_snapshots(config_dir = "config")

# One dataset
list_snapshots("customer_accounts", config_dir = "config")
#>   id  dataset_name          file_name        run_timestamp row_count overall_status
#> 1  1 customer_accounts  delivery_20250130.csv  2025-01-30 09:15:00    150000           PASS
#> 2  4 customer_accounts  delivery_20260128.csv  2026-01-28 10:22:00    158000           WARN
#> 3  7 customer_accounts  delivery_20270129.csv  2027-01-29 08:45:00    162000           FAIL
```

`list_snapshots()` returns a data frame invisibly. `config_dir` tells it
where `dqcheckr.yml` is (to find `snapshot_db`).

### Compare two snapshots

```{r compare-snapshots}
# Latest vs second-latest (default)
drift <- compare_snapshots("customer_accounts", config_dir = "config")

# Specific snapshot IDs — first argument is always treated as "previous"
drift <- compare_snapshots("customer_accounts",
                           snapshot_id_prev = 1,
                           snapshot_id_curr = 7,
                           config_dir = "config")

# Also write a plain-text report alongside the HTML
drift <- compare_snapshots("customer_accounts",
                           snapshot_id_prev = 1,
                           snapshot_id_curr = 7,
                           config_dir  = "config",
                           text_report = TRUE)
```

Console output is a single line:

```
[dqcheckr] drift: customer_accounts snapshot #1 vs #7 | reports/drift_customer_accounts_20270130_091500.html
```

### What the drift report shows

1. **Header** — dataset name, snapshot IDs, timestamps, file names
2. **Table-level drift** — row count change (%), column count, check outcome
   counts (PASS / WARN / FAIL / INFO), threshold breach highlighted
3. **Schema drift** — new columns, dropped columns, type changes
4. **Per-column drift** (sorted by magnitude, breaches highlighted):
   - Missing rate change (percentage points)
   - Non-numeric rate change (percentage points)
   - Numeric mean shift (%)
   - Distinct count change (absolute and %)

### Return value

`compare_snapshots()` returns the full drift data invisibly as a named list,
so you can inspect or process the results programmatically:

```{r drift-programmatic}
drift <- compare_snapshots("customer_accounts",
                           snapshot_id_prev = 1,
                           snapshot_id_curr = 7,
                           config_dir = "config",
                           report = FALSE)   # skip HTML output

# Table-level summary
drift$table_drift

# Schema changes only
drift$schema_changes

# Columns with the largest missing-rate shift
head(drift$missing_rate_changes[, c("Column", "missing_rate_prev",
                                    "missing_rate_curr",
                                    "missing_rate_change_pp",
                                    "missing_rate_exceeds")])

# Columns whose numeric mean shifted most
head(drift$mean_shifts[, c("Column", "numeric_mean_prev",
                            "numeric_mean_curr",
                            "numeric_mean_shift_pct",
                            "numeric_mean_exceeds")])
```

### Typical workflow

```{r drift-workflow}
# 1. Regular delivery run (writes a new snapshot each time)
run_dq_check("customer_accounts", config_dir = "config")

# 2. Two years later, a new delivery arrives
run_dq_check("customer_accounts", config_dir = "config")

# 3. Discover what snapshots are available
list_snapshots("customer_accounts", config_dir = "config")

# 4. Compare any two
compare_snapshots("customer_accounts",
                  snapshot_id_prev = 1,
                  snapshot_id_curr = 4,
                  config_dir = "config")
```

**Note:** When explicit IDs are passed, the function treats the first as
"previous" and the second as "current" regardless of their numeric order.
This matters when baseline snapshots are backfilled from archived files —
their IDs may be higher than earlier production runs.

---

## Worked example — Star Wars dataset

The following uses `dplyr::starwars` exported to CSV and fixed-width files.

```{r starwars-setup}
# Config directory contains:
#   dqcheckr.yml         — global thresholds
#   starwars_csv.yml     — CSV dataset config
#   starwars_fwf.yml     — FWF dataset config

# Run checks on both formats
result_csv <- run_dq_check("starwars_csv", config_dir = "config", open_report = TRUE)
result_fwf <- run_dq_check("starwars_fwf", config_dir = "config", open_report = TRUE)
```

The Star Wars config sets `max_missing_rate: 0.60` because `vehicles` (87%)
and `starships` (77%) are naturally sparse — most characters have none. Both
columns FAIL, which is the correct finding: they are genuinely incomplete and
a consumer should be aware before using them.

Expected console output:

```
[dqcheckr] starwars_csv: FAIL - 0 warning(s), 2 failure(s). Report: output/reports/starwars_csv_....html
[dqcheckr] starwars_fwf: FAIL - 0 warning(s), 2 failure(s). Report: output/reports/starwars_fwf_....html
```

---

## Error handling

| Situation | Behaviour |
|-----------|-----------|
| Config file not found | Stop immediately; no output written |
| Data file not found | Stop immediately; no output written |
| Only one file available | Single-file mode; CP checks skipped |
| File cannot be parsed | Stop with filename and parse error |
| SC-01 / SC-02 violations | FAIL recorded; run continues; report written |
| Custom checks file missing | Stop with message |
| `custom_checks()` not defined | Stop with message |
| `custom_checks()` runtime error | Stop with R error detail |
| SQLite write fails | Warning emitted; HTML report still written |

The package never modifies, moves, or deletes source files.

---

## Design principles

**Owned and evolvable.** Every check is plain R — readable, modifiable, and
not hidden behind a third-party API.

**Plain English outputs.** Reports are written for data management staff, not
engineers. No R code, no package names, no jargon.

**Fail loudly.** Configuration errors, missing files, and custom-check
failures all stop the run with a clear message. Nothing is silently skipped.

**Non-destructive.** The package reads files and writes reports and database
records. Source files are never touched.
