---
title: "dqcheckr — Software Specification"
author: "Mick Mioduszewski"
date: "2026-06-01"
version: "0.2.0"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Software Specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

**Version**: 0.2.0
**Author**: Mick Mioduszewski
**Date**: 2026-06-01

---

## Part A — Purpose and Intent

### Why this exists

The analytics organisation ingests external datasets from multiple suppliers on a recurring basis. Verifying that each new delivery is complete, structurally correct, and consistent with prior deliveries is currently slow and largely manual. Analysts spend time on repetitive visual inspection instead of higher-value work, and quality problems risk being missed or caught late.

### What we are building

`dqcheckr` is a purpose-built R package that automates this verification process. For each dataset arrival it produces a single, plain-English HTML report that tells data management staff:

1. Whether the file is structurally sound (columns, types, row count as expected).
2. Whether data values are within acceptable bounds and free of obvious errors.
3. How this delivery compares to the previous one — what has changed, what has drifted.
4. How this delivery compares to a defined schema contract, if one exists.

Results are also stored in a local SQLite database so that quality trends can be tracked over time across many deliveries.

### Design principles

- **Owned and evolvable.** The tool is first-party R code, not a thin wrapper around a third-party API. Every check is directly readable and modifiable.
- **Plain English outputs.** Reports are written for data management staff, not engineers. No R code, no package names, no jargon.
- **Fail loudly.** Errors in configuration, custom checks, or parsing stop the run with a clear message. Nothing is silently skipped.
- **Non-destructive.** The tool reads files and writes reports and database records. It never modifies, moves, or deletes source files.

---

## Part B — Business-Level Description

### B.1 How it is used

A data officer runs a single command for each arriving dataset:

```{r}
run_dq_check("customer_accounts", config_dir = "config")
```

The tool finds the two most recent files in the dataset's folder, checks the newer one, compares it to the older one, writes a timestamped HTML report to a reports folder, and prints a one-line summary to the console. The report opens automatically in the browser.

Files can also be identified explicitly by naming a current and a previous file, rather than relying on folder modification times.

Scale: the tool is designed for up to 10 dataset arrivals per day across all dataset types.

### B.2 What is checked

**Data completeness**

- Any column with too many missing values is flagged.
- Any column that is entirely empty is an error.
- If the file has fewer rows than the configured minimum, it is an error.

**Data validity**

- Columns expected to contain only specific values (e.g. country codes, status flags) are checked against a permitted list. Unexpected values are an error.
- Numeric columns expected to fall within a range are checked. Out-of-range values are an error.
- Columns expected to match a format (e.g. a postcode pattern) are checked by regular expression. Violations are an error.

**Data integrity**

- Columns designated as keys (e.g. a unique ID column) are checked for duplicates. Duplicates are an error.
- Rows that are entirely identical to another row in the same file are flagged as a warning.

**Data consistency**

- The inferred data type of each column is recorded. A type change between deliveries is flagged.
- Non-numeric values in a column that should be numeric are flagged (warning if any present; error if too many).

**Schema contract** *(when an expected column list is configured)*

- Columns present in the file but not in the expected list are an error.
- Columns in the expected list but absent from the file are an error.

**Custom checks** *(when a custom checks file is configured for a dataset)*

- Organisation-specific rules written in R are run against the file. Results appear alongside the standard checks.

### B.3 What is compared to the previous delivery

- **Row count**: a large change (increase or decrease) is flagged.
- **Schema**: new columns, dropped columns, and type changes are flagged.
- **Missing rates**: columns where the proportion of missing values has increased significantly are flagged.
- **Numeric distributions**: columns where the average value has shifted significantly are flagged.
- **Non-numeric contamination**: columns where the proportion of bad (non-numeric) values has increased are flagged.
- **Distinct values**: new values that have appeared and values that have disappeared in categorical columns are listed.
- **Column order**: a change in the order of columns is a warning for CSV files and an error for fixed-width files (where column order determines parsing).

### B.4 What is produced

**HTML report** — one file per run, self-contained (no internet required to view). Contains:

1. A header showing dataset name, file name, run time, and an overall status badge (green = all clear, amber = warnings, red = failures).
2. A file summary (row count, column count, file size, format, encoding).
3. A quality checks table listing every check, its status, what was observed, the threshold, and a plain-English explanation. Failures and warnings appear first.
4. A custom checks table (if configured).
5. A version comparison section showing what changed since the previous delivery, with a plain-English narrative summary at the top.
6. A historical trend section showing row count and check outcomes across the last 10 runs (appears once at least two runs have been recorded).
7. An appendix with a full column-level statistics table.

**SQLite database** — a single database shared across all datasets. Records summary statistics for every run, enabling long-term trend queries without reading the source files again.

---

## Part C — Technical Specification

### C.1 Package structure

```
dqcheckr/
├── R/
│   ├── run_check.R          # main entry point: run_dq_check()
│   ├── ingest.R             # read CSV / FWF; trim whitespace
│   ├── snapshot.R           # SQLite write + read_recent_snapshots()
│   ├── drift.R              # compare_snapshots(), list_snapshots(), drift helpers
│   ├── checks_generic.R     # QC-01..15 and SC-01..02
│   ├── checks_custom.R      # source and call custom_checks(df)
│   ├── compare.R            # CP-01..08
│   ├── report.R             # quarto::quarto_render() wrapper
│   ├── utils.R              # dq_result(), load_config(), infer_col_type(), helpers
│   └── dqcheckr-package.R   # package-level documentation
├── inst/
│   ├── templates/
│   │   ├── report.qmd       # main HTML report template
│   │   └── drift_report.qmd # snapshot drift report template
│   ├── config/
│   │   ├── dqcheckr.yml
│   │   └── example_dataset.yml
│   └── demonstrations/
│       ├── demo.R           # named-file mode demo (CSV + FWF)
│       ├── demo2.R          # folder-scan mode demo with version comparison
│       ├── README.md
│       ├── config/          # global + starwars_csv + starwars_fwf configs
│       ├── config2/         # global + starwars_folder configs
│       ├── custom2/         # starwars_custom.R (human-specific checks)
│       ├── data/            # starwars.csv, starwars.fwf
│       └── data2/           # starwars_v1.csv (original), starwars_v2.csv (perturbed)
├── tests/testthat/
│   ├── test-utils.R
│   ├── test-checks.R
│   ├── test-compare.R
│   ├── test-ingest.R
│   ├── test-snapshot.R
│   ├── test-drift.R
│   └── test-integration.R
├── vignettes/
│   ├── dqcheckr.Rmd         # user guide
│   └── specification.Rmd    # this document
├── DESCRIPTION
└── NAMESPACE
```

### C.2 Configuration

**Global config — `dqcheckr.yml`**

```yaml
snapshot_db: "data/snapshots.sqlite"
report_output_dir: "reports/"

default_rules:
  # Single-snapshot thresholds
  max_missing_rate: 0.05            # FAIL if column missing rate exceeds this
  max_non_numeric_rate: 0.01        # FAIL if non-numeric rate in numeric column exceeds this
  min_row_count: 0                  # FAIL if row count below this (0 = disabled)
  # max_row_count: 1000000          # optional FAIL if row count exceeds this
  # max_file_size_mb: 500           # optional FAIL if file size exceeds this
  # max_z_score: 3.0                # optional QC-15 outlier detection (Z-score)
  # iqr_fence_multiplier: 1.5       # optional QC-15 outlier detection (IQR fence)

  # Version comparison thresholds
  max_row_count_change_pct: 0.10
  max_numeric_mean_shift_pct: 0.20
  max_missing_rate_change_pp: 2.0   # percentage points
  max_non_numeric_rate_change_pp: 1.0
  missing_rate_change_severity: warn  # "warn" or "fail" — severity for CP-03 breaches

  # Schema flags
  flag_new_columns: true            # set false to suppress new-column warnings in CP-02
  flag_dropped_columns: true        # set false to suppress dropped-column warnings in CP-02
  flag_type_changes: true           # set false to suppress type-change warnings in CP-02
  flag_column_order_change: true    # set false to skip CP-08 entirely
  column_order_severity: warn       # "warn" or "fail" — overrides format default for CP-08
```

**Per-dataset config — `<dataset_name>.yml`**

```yaml
dataset_name: "customer_accounts"
folder: "data/incoming/customer_accounts/"
format: csv                 # csv | fwf
encoding: "UTF-8"
delimiter: ","              # csv only
fwf_widths: ~               # fwf only
fwf_col_names: ~            # fwf only
fwf_skip: 0                 # fwf only — header rows to skip (default 0)

# Optional: explicit file paths override folder-scan version detection
# current_file: "data/incoming/customer_accounts/20260410.csv"
# previous_file: "data/incoming/customer_accounts/20260401.csv"

# Optional: schema contract
expected_columns:
  - id
  - name
  - country_code
  - account_status
  - account_balance
  - created_date

# Optional: key columns that must be unique
key_columns:
  - id

# Optional rule overrides (merged over global defaults)
rule_overrides:
  max_missing_rate: 0.02
  max_row_count_change_pct: 0.05
  max_non_numeric_rate: 0.005
  min_row_count: 1000

# Optional per-column rules
column_rules:
  country_code:
    allowed_values: ["GB", "US", "DE", "FR"]
    pattern: "^[A-Z]{2}$"
  account_status:
    allowed_values: ["ACTIVE", "CLOSED", "SUSPENDED"]
  account_balance:
    min_value: 0
    max_value: 1000000

# Optional custom checks
custom_checks_file: "custom/customer_accounts_checks.R"
```

### C.3 File version detection

Two modes:

1. **Folder scan (default)**: the two most recently modified files in `folder` are used. Newest = current, second-newest = previous.
2. **Explicit naming**: if `current_file` (and optionally `previous_file`) are set in the dataset config, those paths are used directly. Useful when file naming conventions do not reflect delivery order, or when re-running a check for a specific pair of files.

If only a current file is available (either via folder scan or explicit config), the run proceeds in single-file mode: version comparison and vs-previous schema checks are skipped.

### C.4 Ingestion and whitespace trimming

1. File is read entirely as character columns via `readr::read_delim()` or `readr::read_fwf()`.
2. Leading and trailing whitespace is trimmed from every value immediately after reading.
3. The trimmed data frame is used for all subsequent checks and comparisons. The raw file is never modified.

### C.5 Type inference

Applied to each column of the trimmed data frame. Rules applied in order:

| Priority | Condition | Inferred type |
|----------|-----------|---------------|
| 1 | All non-null/non-empty values parse as a known date format | `date` |
| 2 | ≥ 90% of non-null/non-empty values coerce to numeric | `numeric` |
| 3 | Otherwise | `character` |
| — | All values null or empty | `unknown` |

Known date formats tested: `%Y-%m-%d`, `%d/%m/%Y`, `%m/%d/%Y`, `%Y%m%d`, `%d-%m-%Y`.

The 90% threshold means a column with up to 10% bad (non-numeric) values is still classified as numeric, making those bad values visible to QC-11.

### C.6 Quality checks

Each check returns one or more `dq_result` objects:

```{r}
dq_result(check_id, check_name, column = NA, status, observed, threshold = NA, message)
# status: "PASS" | "WARN" | "FAIL" | "INFO"
```

**Single-snapshot checks (QC series)**

| ID | Name | Applies to | FAIL condition | WARN condition |
|----|------|------------|----------------|----------------|
| QC-01 | Missing rate | Every column | `missing_rate > max_missing_rate` | — |
| QC-02 | Empty column | Every column | Column is 100% empty | — |
| QC-03 | Duplicate rows | Table | — | Any fully-duplicate rows exist |
| QC-04 | Row count | Table | — | — (INFO) |
| QC-05 | Column count | Table | — | — (INFO) |
| QC-06 | Inferred type | Every column | — | — (INFO) |
| QC-07 | Numeric stats | Numeric columns | — | — (INFO: min/max/mean/SD) |
| QC-08 | Distinct value count | Character columns | — | — (INFO) |
| QC-09 | Allowed values | Configured columns | Any value outside `allowed_values` | — |
| QC-10 | Numeric bounds | Configured columns | Any value outside `min_value`/`max_value` | — |
| QC-11 | Non-numeric values | Numeric columns | `non_numeric_rate > max_non_numeric_rate` | `non_numeric_rate > warn_non_numeric_rate` |
| QC-12 | Key uniqueness | Configured `key_columns` (single or composite) | Any duplicate value in key | — |
| QC-13 | Pattern / regex | Configured columns | Any value not matching `pattern` | — |
| QC-14 | Row count bounds | Table | `row_count < min_row_count` (if > 0); `row_count > max_row_count` (if set); file size > `max_file_size_mb` (if set) | — |
| QC-15 | Outlier detection | Numeric columns | Value exceeds Z-score (`max_z_score`) or IQR fence (`iqr_fence_multiplier`) threshold | — (skipped if neither threshold is configured) |

**Schema contract checks (SC series)** — only when `expected_columns` is configured

| ID | Name | FAIL condition |
|----|------|----------------|
| SC-01 | Unexpected column | Column in file but not in `expected_columns` |
| SC-02 | Missing expected column | Column in `expected_columns` but absent from file |

**Version comparison checks (CP series)** — skipped in single-file mode

| ID | Name | Flag condition | Severity |
|----|------|----------------|----------|
| CP-01 | Row count change | `abs(pct_change) > max_row_count_change_pct` (WARN if previous has 0 rows) | WARN |
| CP-02a | New columns | Columns in current not in previous (suppressible via `flag_new_columns: false`) | WARN |
| CP-02b | Dropped columns | Columns in previous not in current (suppressible via `flag_dropped_columns: false`) | WARN |
| CP-02c | Column type changes | Type of a common column changed (suppressible via `flag_type_changes: false`) | WARN |
| CP-03 | Missing rate change | Increase > `max_missing_rate_change_pp` pp per column; severity controlled by `missing_rate_change_severity` | WARN (default) / FAIL |
| CP-04 | Numeric mean shift | Shift > `max_numeric_mean_shift_pct` per column | WARN |
| CP-05 | New distinct values | Values in current not in previous (categorical columns) | INFO |
| CP-06 | Dropped distinct values | Values in previous not in current (categorical columns) | INFO |
| CP-07 | Non-numeric rate change | Increase > `max_non_numeric_rate_change_pp` pp per column; always emits a result | WARN |
| CP-08 | Column order change | Column order differs from previous (suppressible via `flag_column_order_change: false`); severity overridable via `column_order_severity` | WARN (CSV) / FAIL (FWF) |

> CP-02 is split into three separate results (CP-02a, CP-02b, CP-02c) so each type of schema change can be independently read in the report and suppressed via config. The `new_cols_vs_previous` and `missing_cols_vs_previous` columns in the `snapshots` table are derived directly from the CP-02a/b results — they are not computed separately.

### C.7 Custom checks

- `custom_checks_file` in dataset config points to an `.R` file.
- Sourced into an isolated environment whose parent is `baseenv()` (not the global env).
- Must define `custom_checks(df)` returning a list of `dq_result` objects.
- `dq_result()` is explicitly injected and available without qualification.
- All other dqcheckr exports (e.g. `resolve_col_type()`) require the `dqcheckr::` prefix.
- Receives the trimmed data frame.
- Errors (missing file, missing function, runtime error) stop the run.
- Results appear in the "Custom Checks" report section and are written to `column_snapshots` (see C.8).

### C.8 SQLite schema

Database and tables created automatically on first run via `init_snapshot_db()`.

**Table: `snapshots`**

| Column | Type | Notes |
|--------|------|-------|
| `id` | INTEGER PK | Auto-increment |
| `dataset_name` | TEXT | |
| `run_timestamp` | TEXT | `YYYY-MM-DD HH:MM:SS` |
| `file_name` | TEXT | Basename of current file |
| `row_count` | INTEGER | |
| `col_count` | INTEGER | |
| `check_pass_count` | INTEGER | |
| `check_warn_count` | INTEGER | |
| `check_fail_count` | INTEGER | |
| `check_info_count` | INTEGER | |
| `overall_status` | TEXT | `FAIL` > `WARN` > `PASS` > `INFO` |
| `new_cols_vs_previous` | TEXT | Comma-separated; NULL if none or no previous |
| `missing_cols_vs_previous` | TEXT | Comma-separated; NULL if none or no previous |
| `new_cols_vs_schema` | TEXT | Comma-separated; NULL if none or `expected_columns` not set |
| `missing_cols_vs_schema` | TEXT | Comma-separated; NULL if none or `expected_columns` not set |
| `comparison_mode` | TEXT | `"comparison"` or `"single"` (single-file runs have no previous) |
| `render_status` | TEXT | `"success"` or `"failed"` (updated if Quarto render fails) |
| `type_changed_cols_vs_previous` | TEXT | Comma-separated `col (prev->curr)` pairs; NULL if none |

**Table: `column_snapshots`**

| Column | Type | Notes |
|--------|------|-------|
| `id` | INTEGER PK | Auto-increment |
| `snapshot_id` | INTEGER FK | → `snapshots.id` |
| `column_name` | TEXT | Column name, or check_id for custom checks |
| `dq_check` | TEXT | Stat name or custom check_id |
| `value` | TEXT | Always stored as text |
| `threshold` | TEXT | NULL if no threshold applies |
| `severity_on_breach` | TEXT | `FAIL`, `WARN`, or NULL if informational |

Rows written per column type:

| `dq_check` | Written for | `threshold` | `severity_on_breach` |
|------------|-------------|-------------|----------------------|
| `inferred_type` | All columns | NULL | NULL |
| `missing_count` | All columns | NULL | NULL |
| `missing_rate` | All columns | `max_missing_rate` | `FAIL` |
| `distinct_count` | All columns | NULL | NULL |
| `numeric_parseable_mean` | Numeric only | NULL | NULL |
| `numeric_sd` | Numeric only | NULL | NULL |
| `numeric_min` | Numeric only | NULL | NULL |
| `numeric_max` | Numeric only | NULL | NULL |
| `non_numeric_count` | Numeric only | NULL | NULL |
| `non_numeric_rate` | Numeric only | `max_non_numeric_rate` | `FAIL` |

Custom check rows: one row per `dq_result` with a non-NA column value. `dq_check` = `check_id`, `value` = `observed`, `threshold` = `threshold`, `severity_on_breach` = status if WARN/FAIL, else NULL.

### C.9 Main entry points

**Standard run:**

```{r}
run_dq_check(
  dataset_name,        # matches <dataset_name>.yml in config_dir
  config_dir = ".",    # contains dqcheckr.yml and <dataset_name>.yml
  open_report = TRUE   # open in browser when interactive
)
# Returns invisibly: list(status, report_path, snapshot_id)
```

Report filename: `<dataset_name>_<YYYYMMDD_HHMMSS>.html`

**Snapshot drift analysis** (compare any two historical snapshots):

```{r}
compare_snapshots(
  dataset_name,
  snapshot_id_prev = NULL,   # defaults to second-most-recent
  snapshot_id_curr = NULL,   # defaults to most-recent
  db_path          = NULL,   # defaults to snapshot_db from dqcheckr.yml
  config_dir       = ".",
  report           = TRUE,   # render HTML drift report
  open_report      = interactive()
)
# Returns invisibly: named list with dataset_name, snap_prev, snap_curr,
#   table_drift, schema_changes, missing_rate_changes, non_numeric_changes,
#   mean_shifts, distinct_changes

list_snapshots(dataset_name = NULL, db_path)
# Returns a data frame of snapshot records; dataset_name = NULL returns all datasets
```

Console output:
```
[dqcheckr] customer_accounts: FAIL - 0 warning(s), 2 failure(s). Report: reports/customer_accounts_20260410_143022.html
```

### C.10 Error handling

| Situation | Behaviour |
|-----------|-----------|
| Folder does not exist | Stop; no output written |
| No files in folder | Stop; no output written |
| Explicit file path not found | Stop; no output written |
| Only one file available | Single-file mode; comparison and vs-previous schema skipped |
| File cannot be parsed | Stop with filename and parse error |
| SC-01 / SC-02 violations | FAIL recorded; run continues; report and snapshot written |
| Custom checks file missing | Stop with message |
| `custom_checks()` not defined | Stop with message |
| `custom_checks()` runtime error | Stop with R error detail |
| SQLite write fails | Warning emitted; HTML report still written |
| Report render fails | Warning emitted; snapshot marked `render_status = "failed"` |
| `snapshot_db` path missing | Created automatically by `init_snapshot_db()` |

### C.11 Dependencies

| Package | Role |
|---------|------|
| `readr` | CSV / FWF ingestion |
| `DBI` + `RSQLite` | Snapshot store |
| `quarto` | Report rendering (Quarto CLI required separately) |
| `knitr` | Template execution during Quarto render |
| `kableExtra` | Styled HTML tables in report templates |
| `ggplot2` + `gridExtra` | Trend charts in report templates |
| `yaml` | Config parsing |
| `dplyr` | Pipe operator (`%>%`) used in report templates |
| `tidyr` | Data reshaping for trend charts in report templates |
| `rlang` | Structured error conditions |

All on CRAN. `pointblank` and `dataCompareR` were evaluated and not used — all logic is implemented natively. `rmarkdown` is listed in `Suggests` for vignette compilation only; it is no longer used for report rendering.

---

## Part D — Unresolved and Deferred Issues

### D.1 Needs design before implementation

| # | Issue | Detail |
|---|-------|--------|
| D-01 | **Explicit file naming — partial spec** | The dataset config allows `current_file` and `previous_file` to override folder-scan detection, but the behaviour when one is set and the other is not has not been fully specified. Implemented as: `current_file` set but `previous_file` absent runs in single-file mode. |
| ~~D-02~~ | ~~**CP-08 column order — WARN vs FAIL severity**~~ | Resolved in v0.2.0: `column_order_severity` config key overrides the format-based default for CP-08. |
| ~~D-03~~ | ~~**QC-03 duplicate rows — no key scope**~~ | Resolved in v0.2.0: QC-12 now supports composite keys via a `key_columns` list. QC-03 retains full-row duplicate detection. |

### D.2 Backlog — agreed

| # | Feature |
|---|---------|
| B-01 | **Cross-column consistency rules** — conditional rules between columns (e.g. if `status = CLOSED` then `close_date` must not be null) |
| B-02 | **Date column recency check** — flag if the most recent value in a date column is older than N days |
| ~~B-03~~ | ~~**Multi-column key uniqueness**~~ | Resolved in v0.2.0: QC-12 now accepts a `key_columns` list for composite keys. |
| B-04 | **`post_run_hook` parameter** — user-supplied function for email/Slack/S3 delivery after a run |
| B-05 | **`batch_run(config_dir)`** — run all configured datasets in sequence |
| B-06 | **Additional input formats** — Parquet (`arrow`), Excel, database tables via `DBI` |
| ~~B-07~~ | ~~**CP-03 severity**~~ | Resolved in v0.2.0: `missing_rate_change_severity` config key controls whether CP-03 breaches are WARN or FAIL. |
| B-08 | **Report themes** — organisation branding via a `theme` parameter |

### D.3 Known limitations

| # | Limitation |
|---|------------|
| L-01 | ~~Type inference uses a fixed 90% threshold~~ — resolved in v0.1.1; configurable via `type_inference_threshold` in `rule_overrides` |
| L-02 | Date formats tested are hardcoded to five common patterns; exotic formats will be misclassified as character |
| L-03 | The report appendix reads from in-memory `col_stats`, not from SQLite. If the SQLite write fails but the report succeeds, the appendix and the database can show slightly different values in edge cases. Report render failures are non-fatal from v0.2.0: the snapshot is preserved and marked `render_status = "failed"`. |
| L-04 | Performance scales linearly with rows but super-linearly with columns. Benchmarked on macOS / aarch64 / R 4.6.0 with 1 M-row synthetic data: 10 cols → 47.6 s (21,031 rows/s); 20 cols → 91.9 s (10,881 rows/s); 49 cols → 138.9 s (7,200 rows/s). Very wide files (100+ columns) will be substantially slower. |
